Tracing the Hercules stream around the Galaxy

### 2010-05-28 tex/hercules.tex
groups-simulations: to do item
groups-simulations: one more to-do item
groups-simulations: add apogee more clearly, some minor edits

%TO DO:
% - l-range of good directions
% - radial and tangential confusion
% - Hercules in Fig 2.
\documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath}
%\usepackage{color,hyperref}
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\figurenames}{\figurename s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\tvector}[1]{\boldsymbol{\vec{#1}}}
\newcommand{\vx}{\tvector{x}}
\newcommand{\vv}{\tvector{v}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\Gaia}{\emph{Gaia}}
\newcommand{\apogee}{APOGEE}
\newcommand{\hipparcos}{\emph{Hipparcos}}
\newcommand{\vR}{\ensuremath{v_R}}
\newcommand{\vphi}{\ensuremath{v_{\phi}}}
\newcommand{\vo}{\ensuremath{v_0}}
\newcommand{\Ro}{\ensuremath{R_0}}
\newcommand{\ro}{\Ro}
\newcommand{\Ab}{\ensuremath{A_b}}
\newcommand{\Rb}{\ensuremath{R_b}}
\newcommand{\Omegab}{\ensuremath{\Omega_b}}
\newcommand{\fdehnen}{\ensuremath{f_{\text{Dehnen}}}}
\newcommand{\sigmaR}{\ensuremath{\sigma_R}}
\newcommand{\rE}{\ensuremath{R_e}}
\newcommand{\Lc}{\ensuremath{L_c}}
\newcommand{\Rs}{\ensuremath{R_s}}
\newcommand{\Rsigma}{\ensuremath{R_{\sigma}}}
\newcommand{\Rolr}{\ensuremath{R_{\text{OLR}}}}

@@

\begin{abstract}
It has been proposed that the Hercules stream, a group of co-moving
stars in the Solar neighborhood offset from the bulk of the velocity
distribution, is the result of resonant interactions between stars in
the outer disk and the Galactic bar. So far it has only been seen in
the immediate Solar neighborhood, but any resonance model make a
prediction over a large fraction of the Galactic disk. I predict the
distribution of stellar velocities and changing Hercules signature as
a function of location in the Galactic disk in a simple model for the
Galaxy and the bar that produces the observed Hercules stream. The
Hercules feature is expected to be strong enough to be unambiguously
detected in the distribution of line-of-sight velocities in selected
directions. I identify quantitatively the most promising lines of
sight for detection in line-of-sight velocities using the
Kullback-Leibler divergence between the predictions from the resonance
model and an axisymmetric model; these directions are all at $l
\gtrsim 250^{\circ}$. The predictions presented here are only weakly
affected by distance uncertainties, assumptions about the distribution
function in the stellar disk, and the details of the Galactic
potential including the effect of spiral structure. \Gaia\ and future
spectroscopic surveys of the Galactic disk will be able to robustly
test the origin of the Hercules stream and constrain the properties of
the Galactic bar.
\end{abstract}

@@

In \figurename~\ref{fig:rphi1d} I show the predicted line-of-sight
velocity distributions in various locations near the Solar circle all
around the Galaxy, obtained by integrating the full two-dimensional
velocity distribution over the tangential velocity. This \figurename\
is no longer symmetric with respect to shifts of 180$^{\circ}$ degree
in azimuth, since the line-of-sight direction breaks this
symmetry. The gray curve is the line-of-sight distribution we would
observe if the targeted stars were in a steady state in an
axisymmetric potential with a Dehnen distribution function
(\eqnname~[\ref{eq:fdehnen}]). Both line-of-sight velocity
distributions are normalized to have equal mass since I assume that
our knowledge of the Galaxy and the tracer population will not be
great enough that we can test predicted differences in the absolute
number of stars at a given location between the axisymmetric and the
bar-resonance predictions. These differences are also largely a
function of the particular form and parameters of the distribution
function that we choose to represent the stellar disk before bar
formation.

The prospects to detect the Hercules feature are quite good on our
side of the Galaxy. The Hercules feature is especially strong in the
line-of-sight velocity distribution for Galactic longitudes $l \geq
250^{\circ}$. This is because of the fortunate coincidence that for
those lines-of-sight we see velocities in the direction along which
the Hercules streams separates most from the bulk of the velocity
distribution (see \figurename~\ref{fig:rphi2d}) and that at these
azimuths the Hercules stream is at the furthest distance from the rest
of the velocity distribution. We see hardly any effect of the bar on
the line-of-sight velocity distributions on the other side of the
Galaxy ($90^{\circ} \leq \phi \leq 270^{\circ}$). On that side we
mostly see the Galactocentric radial velocity in the line-of-sight, in
which the Hercules feature does not separate clearly from the bulk of
the stars. Even though upcoming surveys such as \apogee\ have the
ability to probe the other side of the Galaxy, for the purposes of
detecting the Hercules stream this is uninteresting.

@@

The result of this analysis is shown in
\figurename~\ref{fig:detect}. I present results both using the full
line-of-sight velocity distributions and only considering the part of
the distribution at $|v_{\text{los}}| \geq 0.15 \vo$. As is clear from
\figurename~\ref{fig:obs}, the local velocity distribution contains
many more features than the Hercules stream. These features are mostly
contained to the main mode of the velocity distribution $|v| \lesssim
0.15 \vo$. Observations indicate that these features are also the
result of dynamical interactions \citep[\eg,][]{Bovy10a,sellwood10a},
albeit with spiral arms rather than the bar. Therefore, we can expect
these or similar features to be present in the velocity distribution
at other locations in the Galaxy as well, and thus, if the effect of
the Hercules stream occurs close to the center of the bulk of the
velocity distribution, it might be hard to disentangle from the other
dynamical effects at play. However, these additional dynamical effects
will be much less important for velocities $\gtrsim 0.15
\vo$. \figurename~\ref{fig:detect} shows that there are many
locations, mostly at $l \gtrsim 250^{\circ}$, where the Hercules
stream is clearly offset from the bulk of the velocity distribution
and should be unambiguously detectable.

@@

Since the extinction in the optical bandpasses is rather large in the
directions identified as promising in \figurename~\ref{fig:detect}
\citep[$A_V \approx 8$ mag toward $l \approx
270^{\circ}$;][]{marshall06a}, spectroscopic surveys in the
near-infrared such as \apogee\ are the best option for seeing the
Hercules feature in the line-of-sight direction. The range of Galactic
longitudes that \apogee\ can target along the disk is $-5^{\circ} \leq
l \leq 250^{\circ}$ (SDSS-III Project Description\footnote{Available
at \url{http://sdss3.org/collaboration/description.pdf}~.}). From
\figurename~\ref{fig:detect} this limit appears right on the edge of
where we might hope to detect the Hercules feature. Since \apogee\ is
expected to start taking data in 2011 and to finish operations before
the first \Gaia\ data release, \apogee\ could trace the Hercules
stream around the Galaxy before \Gaia\ does. Since \apogee\ will also
obtain detailed chemical abundances for its target stars, this would
also allow for a more detailed analysis of the abundances of stars in
the Hercules stream \citep[following, \eg,][]{Bensby07a,Bovy10a},
which could shed more light on radial mixing in the Galactic disk.

@@

\clearpage
\begin{figure}
\includegraphics[width=\textwidth]{rphi2d.ps}
\caption{Two-dimensional velocity distributions along the Solar
  circle. The $x$-axis in each subplot is the Galactocentric radial
  velocity and the $y$-axis is the Galactocentric tangential velocity;
  the axis-ranges of each subplot are the same as in
  \figurename~\ref{fig:obs}. The velocity distributions on the other
  side of the Galaxy are obtained by shifting the azimuths by
  180$^{\circ}$.}\label{fig:rphi2d}
\end{figure}

>>>>

\documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath}
%\usepackage{color,hyperref}
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\figurenames}{\figurename s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\tvector}[1]{\boldsymbol{\vec{#1}}}
\newcommand{\vx}{\tvector{x}}
\newcommand{\vv}{\tvector{v}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\Gaia}{\emph{Gaia}}
\newcommand{\apogee}{APOGEE}
\newcommand{\hipparcos}{\emph{Hipparcos}}
\newcommand{\vR}{\ensuremath{v_R}}
\newcommand{\vphi}{\ensuremath{v_{\phi}}}
\newcommand{\vo}{\ensuremath{v_0}}
\newcommand{\Ro}{\ensuremath{R_0}}
\newcommand{\ro}{\Ro}
\newcommand{\Ab}{\ensuremath{A_b}}
\newcommand{\Rb}{\ensuremath{R_b}}
\newcommand{\Omegab}{\ensuremath{\Omega_b}}
\newcommand{\fdehnen}{\ensuremath{f_{\text{Dehnen}}}}
\newcommand{\sigmaR}{\ensuremath{\sigma_R}}
\newcommand{\rE}{\ensuremath{R_e}}
\newcommand{\Lc}{\ensuremath{L_c}}
\newcommand{\Rs}{\ensuremath{R_s}}
\newcommand{\Rsigma}{\ensuremath{R_{\sigma}}}
\newcommand{\Rolr}{\ensuremath{R_{\text{OLR}}}}

@@

\begin{abstract}
It has been proposed that the Hercules stream, a group of co-moving
stars in the Solar neighborhood offset from the bulk of the velocity
distribution, is the result of resonant interactions between stars in
the outer disk and the Galactic bar. So far it has only been seen in
the immediate Solar neighborhood, but any resonance model make a
prediction over a large fraction of the Galactic disk. I predict the
distribution of stellar velocities and changing Hercules signature as
a function of location in the Galactic disk in a simple model for the
Galaxy and the bar that produces the observed Hercules stream. The
Hercules feature is expected to be strong enough to be unambiguously
detected in the distribution of line-of-sight velocities in selected
directions. I identify quantitatively the most promising lines of
sight for detection in line-of-sight velocities using the
Kullback-Leibler divergence between the predictions from the resonance
model and an axisymmetric model; these directions are at $250^{\circ}
\lesssim l \lesssim 290^{\circ}$. The predictions presented here are
only weakly affected by distance uncertainties, assumptions about the
distribution function in the stellar disk, and the details of the
Galactic potential including the effect of spiral structure. \Gaia\
and future spectroscopic surveys of the Galactic disk such as \apogee\
will be able to robustly test the origin of the Hercules stream and
constrain the properties of the Galactic bar.
\end{abstract}

@@

In \figurename~\ref{fig:rphi1d} I show the predicted line-of-sight
velocity distributions in various locations near the Solar circle all
around the Galaxy, obtained by integrating the full two-dimensional
velocity distribution over the velocity perpendicular to the line of
sight. This \figurename\ is no longer symmetric with respect to shifts
of 180$^{\circ}$ degree in azimuth, since the line-of-sight direction
breaks this symmetry. The gray curve is the line-of-sight distribution
we would observe if the targeted stars were in a steady state in an
axisymmetric potential with a Dehnen distribution function
(\eqnname~[\ref{eq:fdehnen}]). Both line-of-sight velocity
distributions are normalized to have equal mass since I assume that
our knowledge of the Galaxy and the tracer population will not be
great enough that we can test predicted differences in the absolute
number of stars at a given location between the axisymmetric and the
bar-resonance predictions. These differences are also largely a
function of the particular form and parameters of the distribution
function that we choose to represent the stellar disk before bar
formation.

The prospects to detect the Hercules feature are quite good on our
side of the Galaxy. The Hercules feature is especially strong in the
line-of-sight velocity distribution for Galactic longitudes
$250^{\circ} \lesssim l \lesssim 290^{\circ}$. This is because of the
fortunate coincidence that for those lines-of-sight we see velocities
in the direction along which the Hercules streams separates most from
the bulk of the velocity distribution (see
\figurename~\ref{fig:rphi2d}) and that at these azimuths the Hercules
stream is at the furthest distance from the rest of the velocity
distribution. We see hardly any effect of the bar on the line-of-sight
velocity distributions on the other side of the Galaxy ($90^{\circ}
\leq \phi \leq 270^{\circ}$). On that side we mostly see the
Galactocentric radial velocity in the line-of-sight, in which the
Hercules feature does not separate clearly from the bulk of the
stars. Even though upcoming surveys such as \apogee\ have the ability
to probe the other side of the Galaxy, for the purposes of detecting
the Hercules stream this is uninteresting.

@@

The result of this analysis is shown in
\figurename~\ref{fig:detect}. I present results both using the full
line-of-sight velocity distributions and only considering the part of
the distribution at $|v_{\text{los}}| \geq 0.15 \vo$. As is clear from
\figurename~\ref{fig:obs}, the local velocity distribution contains
many more features than the Hercules stream. These features are mostly
contained to the main mode of the velocity distribution $|v| \lesssim
0.15 \vo$. Observations indicate that these features are also the
result of dynamical interactions \citep[\eg,][]{Bovy10a,sellwood10a},
albeit with spiral arms rather than the bar. Therefore, we can expect
these or similar features to be present in the velocity distribution
at other locations in the Galaxy as well, and thus, if the effect of
the Hercules stream occurs close to the center of the bulk of the
velocity distribution, it might be hard to disentangle from the other
dynamical effects at play. However, these additional dynamical effects
will be much less important for velocities $\gtrsim 0.15
\vo$. \figurename~\ref{fig:detect} shows that there are many
locations, mostly at $250^{\circ} \lesssim l \lesssim 290^{\circ}$,
where the Hercules stream is clearly offset from the bulk of the
velocity distribution and should be unambiguously detectable.

@@

Since the extinction in the optical bandpasses is rather large in the
directions identified as promising in \figurename~\ref{fig:detect}
\citep[$A_V \approx 8$ mag toward $l \approx
270^{\circ}$;][]{marshall06a}, spectroscopic surveys in the
near-infrared such as \apogee\ are the best option for seeing the
Hercules feature in the line-of-sight direction. The range of Galactic
longitudes that \apogee\ can target along the disk is $-5^{\circ} \leq
l \leq 250^{\circ}$ (SDSS-III Project Description\footnote{Available
at \url{http://sdss3.org/collaboration/description.pdf}~.}). From
\figurename~\ref{fig:detect} this limit is right on the edge of where
we might hope to detect the Hercules feature. The hercules bump in the
direction $l = 250^{\circ}$ is about half as strong as in
\figurename~\ref{fig:1dvar}, but remains as a clear signature of the
bar. Since \apogee\ is expected to start taking data in 2011 and to
finish operations before the first \Gaia\ data release, \apogee\ could
trace the Hercules stream around the Galaxy before \Gaia\ does. Since
\apogee\ will also obtain detailed chemical abundances for its target
stars, this would also allow for a more detailed analysis of the
abundances of stars in the Hercules stream \citep[following,
\eg,][]{Bensby07a,Bovy10a}, which could shed more light on radial
mixing in the Galactic disk.

@@

\clearpage
\begin{figure}
\includegraphics[width=\textwidth]{rphi2d.ps}
\caption{Two-dimensional velocity distributions along the Solar circle
in the bar-resonance model for the Hercules stream. The $x$-axis in
each subplot is the Galactocentric radial velocity and the $y$-axis is
the Galactocentric tangential velocity; the axis-ranges of each
subplot are the same as in \figurename~\ref{fig:obs}. The velocity
distributions on the other side of the Galaxy are obtained by shifting
the azimuths by 180$^{\circ}$.}\label{fig:rphi2d}
\end{figure}
